### 2015-06-18 doc/thesis/system.tex
Back porting the section about subtyping from paper to thesis

We present the subtyping rules as a deduction system for the
subtyping relation $\senv \vdash T_{1} \subtype T_{2}$.
The variable $\senv$ is a set of pairs of recursion variables.
We need this set to record the hypotheses that we assume when checking
recursive types.

@@

Our type system includes the top type $\Value$,
so any first-level type is a subtype of $\Value$:
\[
\begin{array}{c}
\mylabel{S-VALUE}\\
\senv \vdash T \subtype \Value
\end{array}
\]

@@

The subtyping rules for union types are standard:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNION1}\\
\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
       \senv \vdash T_{2} \subtype T}
      {\senv \vdash T_{1} \cup T_{2} \subtype T}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION2}\\
\dfrac{\senv \vdash T \subtype T_{1}}
      {\senv \vdash T \subtype T_{1} \cup T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION3}\\
\dfrac{\senv \vdash T \subtype T_{2}}
      {\senv \vdash T \subtype T_{1} \cup T_{2}}
\end{array}
\end{array}
\]

The first rule shows that a union type $T_{1} \cup T_{2}$
is a subtype of $T$ if both $T_{1}$ and $T_{2}$ are subtypes
of $T$;
and the other rules show that a type $T$ is a subtype
of a union type $T_{1} \cup T_{2}$ if $T$ is a subtype of
either $T_{1}$ or $T_{2}$.

@@

The type $\Void$ is a subtype of itself and of a variadic type:
\[
\begin{array}{c}
\mylabel{S-VOID2}\\
\senv \vdash \Void \subtype T{*}
\end{array}
\]

A variadic type $T{*}$ represents a sequence of values of type
$T \cup \Nil$, and the rule \textsc{S-VOID2} handles the case where
a given sequence is empty.

The subtyping rule for pair types is the standard covariant rule:
\[
\begin{array}{c}
\mylabel{S-PAIR}\\
\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
       \senv \vdash P_{1} \subtype P_{2}}
      {\senv \vdash T_{1} \times P_{1} \subtype T_{2} \times P_{2}}
\end{array}
\]

The subtyping rules for variadic types are not so obvious.
We need six different subtyping rules for variadic types
to handle all the cases where they can appear.

The rule \textsc{S-VARARG1} is a special rule for handling the
case where we give a sequence of $\Nil$ to the empty tuple:
\[
\begin{array}{c}
\mylabel{S-VARARG1}\\
\senv \vdash \Nil{*} \subtype \Void
\end{array}
\]

The rule \textsc{S-VARARG2} handles the case where both tuple types end
with variadic types:
\[
\begin{array}{c}
\mylabel{S-VARARG2}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
      {\senv \vdash T_{1}{*} \subtype T_{2}{*}}
\end{array}
\]

This rule shows that $T_{1}{*}$ is a subtype of $T_{2}{*}$
if $T_{1} \cup \Nil$ is a subtype of $T_{2} \cup \Nil$.
It explicitly includes $\Nil$ in both sides because otherwise
$\Nil{*}$ would not be a subtype of several other variadic types.
For instance, $\Nil{*}$ would not be a subtype of $\Number{*}$,
as $\Nil \not\subtype \Number$.

The other rules handle the cases where only one tuple type ends with a variadic type:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VARARG3}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2}}
      {\senv \vdash T_{1}{*} \subtype T_{2} \times \Void}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG4}\\
\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil}
      {\senv \vdash T_{1} \times \Void \subtype T_{2}{*}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VARARG5}\\
\dfrac{\senv \vdash T_{1}{*} \subtype T_{2} \times \Void \;\;\;
       \senv \vdash T_{1}{*} \subtype P_{2}}
      {\senv \vdash T_{1}{*} \subtype T_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VARARG6}\\
\dfrac{\senv \vdash T_{1} \times \Void \subtype T_{2}{*} \;\;\;
       \senv \vdash P_{1} \subtype T_{2}{*}}
      {\senv \vdash T_{1} \times P_{1} \subtype T_{2}{*}}
\end{array}
\end{array}
\]

Note that the case where both tuple types end with the type $\Void$ does
not require any special rule.
In the next section we will show that we use the subtyping rules for variadic types,
along with the types $\Value$ and $\Nil$, to make our type system reflect
the semantics of Lua on discarding extra parameters and
replacing missing parameters.

@@

Back to the subtyping rules between first-level types,
the subtyping rule among a \emph{fixed} or \emph{closed}
table type and another \emph{closed} table type resembles the
standard subtyping rule between records:
\[
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
       \senv \vdash V_{j} \subtype_{c} V_{i}'}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}} \; m \ge n
\end{array}
\]

The rule \textsc{S-TABLE1} allows width subtyping and introduces the
auxiliary relation $\subtype_{c}$ to handle depth subtyping on the
type of the values stored in the table fields.
We need an auxiliary relation because the subtyping of the
type of the values stored in the table fields changes according to
the tags of the table types.
We define the relation $\subtype_{c}$ as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
       \senv \vdash V_{2} \subtype V_{1}}
      {\senv \vdash V_{1} \subtype_{c} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\end{array}
\]

@@

We introduced \emph{unique} table types to avoid this limitation,
as they represent the type of tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, this is the case of the table constructor.
The following subtyping rule defines the subtyping relation among
\emph{unique} table types and \emph{closed} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}
\]

@@

The rule \textsc{S-TABLE2} introduced the auxiliary relations
$\subtype_{u}$ and $\subtype_{o}$.
The first allows depth subtyping on all fields,
while the second allows the omission of optional fields.
We define them as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD4}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{u} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD5}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD6}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD7}\\
\dfrac{\senv \vdash \Nil \subtype V}
      {\senv \vdash \Nil \subtype_{o} V}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD8}\\
\dfrac{\senv \vdash \Nil \subtype V}
      {\senv \vdash \Nil \subtype_{o} \Const \; V}
\end{array}
\end{array}
\]

@@

Even though we allow width subtyping between \emph{unique} and \emph{closed}
table types, we do not allow it among \emph{unique} and other table types
because it would violate our definition of these other table types:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}}
\end{array}
\]

The rule that handles subtyping between \emph{open} and \emph{closed} table
types allows width subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE4}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}
\]

However, the rule that handles subtyping among \emph{open} and
\emph{open} or \emph{fixed} table types does not allow width subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}}
\end{array}
\]

@@

We introduced \emph{fixed} table types because we needed a safe way
to represent the type of classes that can allow single inheritance
through the refinement of table types.
The rule that handles subtyping between \emph{fixed} table types
does not allow width subtyping, joining fields, and omitting fields,
but it allows depth subtyping on $\Const$ fields:
\[
\begin{array}{c}
\mylabel{S-TABLE6}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..n \;\;\;
       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
       \senv \vdash V_{j} \subtype_{c} V_{i}'}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{fixed}}
\end{array}
\]

@@

We use the \emph{Amber rule} \cite{cardelli1986amber} to define
subtyping between recursive types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-AMBER}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
      {\senv \vdash \mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-ASSUMPTION}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
      {\senv \vdash x_{1} \subtype x_{2}}
\end{array}
\end{array}
\]

The rule \textsc{S-AMBER} also uses the rule \textsc{S-ASSUMPTION}
to check whether $\mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}$.
Both rules use the set of assumptions $\senv$,
where each assumption is a pair of recursion variables.
The rule \textsc{S-AMBER} extends $\senv$ with the assumption
$x_{1} \subtype x_{2}$ to check whether $T_{1} \subtype T_{2}$.
The rule \textsc{S-ASSUMPTION} allows the rule \textsc{S-AMBER}
to check whether an assumption is valid.

A recursive type may appear inside a first-level type, and our
type system includes subtyping rules to handle subtyping between
recursive types and other first-level types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNFOLDR}\\
\dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
      {\senv \vdash T_{1} \subtype \mu x.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNFOLDL}\\
\dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
      {\senv \vdash \mu x.T_{1} \subtype T_{2}}
\end{array}
\end{array}
\]

@@

Filter types are subtypes only of themselves and of $\Value$.
More precisely, a filter type $\phi(T_{1},T_{2})$ is a subtype of
the same filter type $\phi(T_{1},T_{2})$, which shares the same
types $T_{1}$ and $T_{2}$, and it is also a subtype of $\Value$.

@@

Even though the dynamic type $\Any$ does not interact with subtyping,
it does interact with consistent-subtyping.
We present the consistent-subtyping rules as a deduction system for
the consistent-subtyping relation $\senv \vdash T_{1} \lesssim T_{2}$.
As in the subtyping relation, $\senv$ is also a set of pairs of
recursion variables.
We define the consistent-subtyping rules for the dynamic type $\Any$
as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{C-ANY1}\\
\senv \vdash T \lesssim \Any
\end{array}
\;
\begin{array}{c}
\mylabel{C-ANY2}\\
\senv \vdash \Any \lesssim T
\end{array}
\end{array}
\]

If we had set the type $\Any$ as both bottom and top types of our
subtyping relation, then any type $T_{1}$ would be a subtype of
any other type $T_{2}$.
The consequence of this is that all programs would type check without errors.
This would happen due to the transitivity of subtyping, that is,
we would be able to down-cast any type $T_{1}$ to $\Any$ and then up-cast
$\Any$ to any other type $T_{2}$.
The rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that
allow the dynamic type to interact with other first-level types,
and thus allow dynamically typed code to coexist with statically
typed code.
Because of these two rules, consistent-subtyping cannot be transitive.
These two rules are the only rules that differ between
subtyping and consistent-subtyping, if we implement the subtyping rules
as we do in this section.

>>>>

We present the subtyping rules as a deduction system for the
subtyping relation $\senv \vdash X \subtype Y$, where $X$ and $Y$
are two types of the same kind.
The variable $\senv$ is a set of pairs of recursion variables.
We need this set to record the hypotheses that we assume when checking
recursive types.

@@

Our type system includes the top type $\Value$,
so any first-level type is a subtype of $\Value$:
\[
\begin{array}{c}
\mylabel{S-VALUE}\\
\senv \vdash F \subtype \Value
\end{array}
\]

@@

The subtyping rules for union types are standard:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNION1}\\
\dfrac{\senv \vdash F_{1} \subtype F \;\;\;
       \senv \vdash F_{2} \subtype F}
      {\senv \vdash F_{1} \cup F_{2} \subtype F}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION2}\\
\dfrac{\senv \vdash F \subtype F_{1}}
      {\senv \vdash F \subtype F_{1} \cup F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION3}\\
\dfrac{\senv \vdash F \subtype F_{2}}
      {\senv \vdash F \subtype F_{1} \cup F_{2}}
\end{array}
\end{array}
\]

The first rule shows that a union type $F_{1} \cup F_{2}$
is a subtype of $F$ if both $F_{1}$ and $F_{2}$ are subtypes
of $F$;
and the other rules show that a type $F$ is a subtype
of a union type $F_{1} \cup F_{2}$ if $F$ is a subtype of
either $F_{1}$ or $F_{2}$.

@@

The subtyping rule for pair types is the standard covariant rule:
\[
\begin{array}{c}
\mylabel{S-PAIR}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
       \senv \vdash P_{1} \subtype P_{2}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\]

The subtyping rules for variadic types are not so obvious.
We need three different subtyping rules for variadic types
to handle all the cases where they can appear.

The rule \textsc{S-VARARG1} handles the case where both tuple types end
with variadic types:
\[
\begin{array}{c}
\mylabel{S-VARARG2}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
      {\senv \vdash F_{1}{*} \subtype F_{2}{*}}
\end{array}
\]

This rule shows that $F_{1}{*}$ is a subtype of $F_{2}{*}$
if $F_{1} \cup \Nil$ is a subtype of $F_{2} \cup \Nil$.
It explicitly includes $\Nil$ in both sides because otherwise
$\Nil{*}$ would not be a subtype of several other variadic types.
For instance, $\Nil{*}$ would not be a subtype of $\Number{*}$,
as $\Nil \not\subtype \Number$.

The other rules handle the cases where only one tuple type ends with a variadic type:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VARARG2}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
       \senv \vdash F_{1}{*} \subtype P_{2}}
      {\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG3}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
       \senv \vdash P_{1} \subtype F_{2}{*}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
\end{array}
\end{array}
\]

These rules state the conditions when variadic tuples of different
length are compatible.
In the next section we will show that we use the subtyping rules for variadic types,
along with the types $\Value$ and $\Nil$, to make our type system reflect
the semantics of Lua on discarding extra parameters and
replacing missing parameters.

@@

Back to the subtyping rules between first-level types,
the subtyping rule among a \emph{fixed} or \emph{closed}
table type and another \emph{closed} table type resembles the
standard subtyping rule between records:
\[
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \; \exists j \;\;\;
       \senv \vdash F_{j} \subtype F_{i}' \;\;\;
       \senv \vdash F_{i}' \subtype F_{j} \;\;\;
       \senv \vdash V_{j} \subtype_{c} V_{i}'}
      {\senv \vdash \{\overline{F{:}V}\}_{fixed|closed} \subtype
                    \{\overline{F'{:}V'}\}_{closed}}
\end{array}
\]

The rule \textsc{S-TABLE1} allows width subtyping and introduces the
auxiliary relation $\subtype_{c}$ to handle depth subtyping on the
type of the values stored in the table fields.
We need an auxiliary relation because the subtyping of the
type of the values stored in the table fields changes according to
the tags of the table types.
We define the relation $\subtype_{c}$ as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
       \senv \vdash F_{2} \subtype F_{1}}
      {\senv \vdash F_{1} \subtype_{c} F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
      {\senv \vdash \Const \; F_{1} \subtype_{c} \Const \; F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
      {\senv \vdash F_{1} \subtype_{c} \Const \; F_{2}}
\end{array}
\end{array}
\]

@@

We introduced \emph{unique} table types to avoid this limitation,
as they represent the type of tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, this is the case of the table constructor.
The following subtyping rule defines the subtyping relation among
\emph{unique} table types and \emph{closed} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\begin{array}{c}
       \forall i \; \forall j \;\;\;
       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}'\\
       \forall j \; \nexists i \;\;\;
       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{\overline{F{:}V}\}_{unique} \subtype
                    \{\overline{F'{:}V'}\}_{closed}}
\end{array}
\]

@@

The rule \textsc{S-TABLE2} introduced the auxiliary relations
$\subtype_{u}$ and $\subtype_{o}$.
The first allows depth subtyping on all fields,
while the second allows the omission of optional fields.
We define them as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD4}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
      {\senv \vdash F_{1} \subtype_{u} F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD5}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
      {\senv \vdash \Const \; F_{1} \subtype_{u} \Const \; F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD6}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
      {\senv \vdash \Const \; F_{1} \subtype_{u} F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD7}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
      {\senv \vdash F_{1} \subtype_{u} \Const \; F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD8}\\
\dfrac{\senv \vdash \Nil \subtype F}
      {\senv \vdash \Nil \subtype_{o} F}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD9}\\
\dfrac{\senv \vdash \Nil \subtype F}
      {\senv \vdash \Nil \subtype_{o} \Const \; F}
\end{array}
\end{array}
\]

@@

Even though we allow width subtyping between \emph{unique} and \emph{closed}
table types, we do not allow it among \emph{unique} and other table types
because it would violate our definition of these other table types:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\begin{array}{c}
       \forall i \; \exists j \;\;\;
       \senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \; \nexists i \;\;\;
       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{\overline{F{:}V}\}_{unique} \subtype
                    \{\overline{F'{:}V'}\}_{unique|open|fixed}}
\end{array}
\]

The rule that handles subtyping between \emph{open} and \emph{closed} table
types allows width subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE4}\\
\dfrac{\begin{array}{c}
       \forall i \; \forall j \;\;\;
       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \forall j \; \nexists i \;\;\;
       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{\overline{F{:}V}\}_{open} \subtype
                    \{\overline{F'{:}V'}\}_{closed}}
\end{array}
\]

However, the rule that handles subtyping among \emph{open} and
\emph{open} or \emph{fixed} table types does not allow width subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\begin{array}{c}
       \forall i \; \exists j \;\;\;
       \senv \vdash F_{i} \subtype F_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \forall j \; \nexists i \;\;\;
       \senv \vdash F_{i} \subtype F_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{\overline{F{:}V}\}_{open} \subtype
                    \{\overline{F'{:}V'}\}_{open|fixed}}
\end{array}
\]

@@

We introduced \emph{fixed} table types because we needed a safe way
to represent the type of classes that can allow single inheritance
through the refinement of table types.
The rule that handles subtyping between \emph{fixed} table types
does not allow width subtyping, joining fields, and omitting fields,
but it allows depth subtyping on $\Const$ fields:
\[
\begin{array}{c}
\mylabel{S-TABLE6}\\
\dfrac{\begin{array}{c}
       \forall i \; \exists j \;\;\;
       \senv \vdash F_{i} \subtype F_{j}' \;\;\;
       \senv \vdash F_{j}' \subtype F_{i} \;\;\;
       \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \forall j \; \exists i \;\;\;
       \senv \vdash F_{i} \subtype F_{j}' \;\;\;
       \senv \vdash F_{j}' \subtype F_{i} \;\;\;
       \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \end{array}}
      {\senv \vdash \{\overline{F{:}V}\}_{fixed} \subtype
                    \{\overline{F'{:}V'}\}_{fixed}}
\end{array}
\]

@@

We use the \emph{Amber rule} \cite{cardelli1986amber} to define
subtyping between recursive types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-AMBER}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash F_{1} \subtype F_{2}}
      {\senv \vdash \mu x_{1}.F_{1} \subtype \mu x_{2}.F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-ASSUMPTION}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
      {\senv \vdash x_{1} \subtype x_{2}}
\end{array}
\end{array}
\]

The rule \textsc{S-AMBER} also uses the rule \textsc{S-ASSUMPTION}
to check whether $\mu x_{1}.F_{1} \subtype \mu x_{2}.F_{2}$.
Both rules use the set of assumptions $\senv$,
where each assumption is a pair of recursion variables.
The rule \textsc{S-AMBER} extends $\senv$ with the assumption
$x_{1} \subtype x_{2}$ to check whether $F_{1} \subtype F_{2}$.
The rule \textsc{S-ASSUMPTION} allows the rule \textsc{S-AMBER}
to check whether an assumption is valid.

A recursive type may appear inside a first-level type, and our
type system includes subtyping rules to handle subtyping between
recursive types and other first-level types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNFOLDR}\\
\dfrac{\senv \vdash F_{1} \subtype [x \mapsto \mu x.F_{2}]F_{2}}
      {\senv \vdash F_{1} \subtype \mu x.F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNFOLDL}\\
\dfrac{\senv \vdash [x \mapsto \mu x.F_{1}]F_{1} \subtype F_{2}}
      {\senv \vdash \mu x.F_{1} \subtype F_{2}}
\end{array}
\end{array}
\]

@@

Filter types are subtypes only of themselves and of $\Value$.
More precisely, a filter type $\phi(F_{1},F_{2})$ is a subtype of
the same filter type $\phi(F_{1},F_{2})$, which shares the same
types $F_{1}$ and $F_{2}$, and it is also a subtype of $\Value$.

@@

Even though the dynamic type $\Any$ does not interact with subtyping,
it does interact with consistent-subtyping.
We present the consistent-subtyping rules as a deduction system for
the consistent-subtyping relation $\senv \vdash X \lesssim Y$,
where $X$ and $Y$ are two types of the same kind.
As in the subtyping relation, $\senv$ is also a set of pairs of
recursion variables.
We define the consistent-subtyping rules for the dynamic type $\Any$
as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{C-ANY1}\\
\senv \vdash F \lesssim \Any
\end{array}
\;
\begin{array}{c}
\mylabel{C-ANY2}\\
\senv \vdash \Any \lesssim F
\end{array}
\end{array}
\]

If we had set the type $\Any$ as both bottom and top types of our
subtyping relation, then any type $F_{1}$ would be a subtype of
any other type $F_{2}$.
The consequence of this is that all programs would type check without errors.
This would happen due to the transitivity of subtyping, that is,
we would be able to down-cast any type $F_{1}$ to $\Any$ and then up-cast
$\Any$ to any other type $F_{2}$.
The rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that
allow the dynamic type to interact with other first-level types,
and thus allow dynamically typed code to coexist with statically
typed code.
Because of these two rules, consistent-subtyping cannot be transitive.
These two rules are the only rules that differ between
subtyping and consistent-subtyping, if we implement the subtyping rules
as we do in this section.
